  ---
title: "Lecture 3: BACI designs and Linear models"
output:
  html_document:
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
---

# ENVX3002 Topic 2: ANOVA and its extensions - BACI design

## Generate data for a BACI anova design
```{r}

# Load necessary library
library(dplyr)

# Set seed for reproducibility
set.seed(123)

# Define parameters
n <- 50  # number of observations per group
mu_before_control <- 10  # mean response in control site before intervention
mu_before_impact <- 10  # mean response in impact site before intervention
mu_after_control <- 10  # mean response in control site after intervention
mu_after_impact <- 15  # mean response in impact site after intervention (effect of intervention)
sd <- 2  # standard deviation of response

# Simulate data and set level order for the factors
data <- data.frame(
  Site = rep(c("Control", "Impact"), each = 2*n),
  Time = rep(c("Before", "After"), each = n, times = 2),
  Response = c(
    rnorm(n, mu_before_control, sd),
    rnorm(n, mu_after_control, sd),
    rnorm(n, mu_before_impact, sd),
    rnorm(n, mu_after_impact, sd)
  )
) %>% 
  mutate(Site = factor(Site, levels = c("Control", "Impact")),
         Time = factor(Time, levels = c("Before", "After")))


# Print the first few rows of the data
head(data)

```


## Anova analysis of BACI design
```{r}
model <- aov(Response ~ Site*Time, data = data)

# Print the summary of the model
summary(model)

```

## Post-hoc test
```{r}
library(emmeans)
emmeans(model, pairwise ~ Site:Time)

```


## Plotting results
```{r}

emmip(model, Site ~ Time, CIs = TRUE)

```

## Test assumptions of ANOVA
```{r}

par(mfrow=c(2,2))
plot(model)
par(mfrow=c(1,1))

```


```{r}
# Test assumptions of ANOVA using performance package
library(performance)
check_model(model)
```


# Beyond BACI design

## Generate code for a Beyond BACI design: add two control sites and one impact site

```{r}
# Load necessary library
library(dplyr)

# Set seed for reproducibility
set.seed(123)

# Define parameters
n <- 50  # number of observations per group
mu_before_control <- 10  # mean response in control site before intervention
mu_before_impact <- 10  # mean response in impact site before intervention

mu_after_control <- 10  # mean response in control site after intervention
mu_after_impact <- 15  # mean response in impact site after intervention (effect of intervention)

sd <- 2  # standard deviation of response

# Simulate data and set level order for the factors
data <- data.frame(
  Site = rep(c("Control1", "Control2", "Impact"), each = 2*n),
  Time = rep(c("Before", "After"), each = n, times = 3),
  Response = c(
    rnorm(n, mu_before_control, sd),
    rnorm(n, mu_after_control, sd),
    rnorm(n, mu_before_control, sd),
    rnorm(n, mu_after_control, sd),
    rnorm(n, mu_before_impact, sd),
    rnorm(n, mu_after_impact, sd)
  )
) %>% 
  mutate(Site = factor(Site, levels = c("Control1", "Control2", "Impact")),
         Time = factor(Time, levels = c("Before", "After")))
# Print the first few rows of the data
head(data)



```


## Anova analysis of Beyond BACI design
```{r}
model <- aov(Response ~ Site*Time, data = data)

# Print the summary of the model
summary(model)

```

## Post-hoc test
```{r}

emmeans(model, pairwise ~ Site:Time)

```

## Plotting results
```{r}

emmip(model, Site ~ Time, CIs = TRUE)

```

## Test assumptions of ANOVA
```{r}

par(mfrow=c(2,2))
plot(model)
par(mfrow=c(1,1))

```

```{r}
# Test assumptions of ANOVA using performance package
check_model(model)
```



# ENVX3002 Topic 3: General linear model - R code for Lectures

## Analysis of chick weights via regression on indicator variables

```{r}
chicks <- read.csv("Chick weights.csv")
str(chicks)
chicks$Diet <- as.factor(chicks$Diet)
str(chicks)
```

```{r}
aggregate(WtGain ~ Diet, mean, data = chicks)
```

```{r}
chicks.aov <- aov(WtGain ~ Diet, data = chicks)
summary(chicks.aov)
```

### Fit model using indicator variables

```{r}
d1 <- 1 * (chicks$Diet == 1)
d2 <- 1 * (chicks$Diet == 2)
d3 <- 1 * (chicks$Diet == 3)
d4 <- 1 * (chicks$Diet == 4)
data.frame(chicks$Diet, d1, d2, d3, d4)
```

```{r}
summary(lm(WtGain ~ d1 + d2 + d3 + d4, data = chicks))
```

```{r}
chicks.refl.lm <- lm(WtGain ~ d2 + d3 + d4, data = chicks)
summary(chicks.refl.lm)
```

```{r}
chicks.ref4.lm <- lm(WtGain ~ d1 + d2 + d3, data = chicks)
summary(chicks.ref4.lm)
```

### Fit model using Diet as categorical varable

```{r}
chicks.lm <- lm(WtGain ~ Diet, data = chicks)
summary(chicks.lm)
```

## Analysis of sheep weaning weights via a general linear model

```{r}
sheep <- read.csv("Kenya sheep.csv")
str(sheep)
sheep$Breed <- as.factor(sheep$Breed)
str(sheep)
```

```{r}
summary(sheep)
```

```{r}
plot(WeanWt ~ BirthWt, data = sheep)
```

```{r}
plot(WeanWt ~ BirthWt, xlab = "Birth weight (kg)", ylab = "Weaning weight (kg)",
   col = 1:4, cex = 0.5, data = sheep)
legend("topleft", levels(sheep$Breed), pch = 1, col = 1:4) # note we changed Breed to be a factor above to take advantage of plotting levels as different colours

```

### Fit main effects model

```{r}
sheep.lm1 <- lm(WeanWt ~ BirthWt + Breed, data = sheep)
summary(sheep.lm1)
```

```{r}
drop1(sheep.lm1, test = "F")

```

```{r}
anova(sheep.lm1, test = "F")
```

```{r}
par(mfrow=c(1,2))
plot(sheep.lm1, which = 1)
plot(sheep.lm1, which = 2)
par(mfrow=c(1,2))
```

### Fit model with interaction

```{r}
sheep.lm2 <- lm(WeanWt ~ BirthWt * Breed, data = sheep)
summary(sheep.lm2)
```

```{r}
anova(sheep.lm2, test = "F")
```

### Plot fitted model - no interaction

```{r}
(sheep.pred.df <- with(sheep,
   expand.grid(BirthWt = c(min(BirthWt), max(BirthWt)),
               Breed = levels(Breed))))
```

```{r}

WeanWt.pred <- predict(sheep.lm1, sheep.pred.df)

(sheep.pred.df <- cbind(sheep.pred.df, WeanWt.pred))
```

```{r}
plot(WeanWt ~ BirthWt, xlab = "Birth weight (kg)", ylab = "Weaning weight (kg)",
   col = Breed, cex = 0.5, data = sheep)
legend("topleft", levels(sheep$Breed), pch = 1, col = 1:4)

lines(sheep.pred.df[1:2, -2], col = 1) # Breed 1, DD
lines(sheep.pred.df[3:4, -2], col = 2) # Breed 2, DR
lines(sheep.pred.df[5:6, -2], col = 3) # Breed 3, RD
lines(sheep.pred.df[7:8, -2], col = 4) # Breed 4, RR



mean.WW <- mean(sheep$WeanWt, na.rm = TRUE)
mean.BW <- mean(sheep$BirthWt, na.rm = TRUE)

abline(h = mean.WW, lty = 2, col = "grey")
abline(v = mean.BW, lty = 2, col = "grey")

```

### Plot fitted model - with interaction (for illustration only)

```{r}
plot(WeanWt ~ BirthWt, xlab = "Birth weight (kg)", ylab = "Weaning weight (kg)",
   col = Breed, cex = 0.5, data = sheep)
legend("topleft", levels(sheep$Breed), pch = 1, col = 1:4)

sheep.pred.df <- with(sheep,
   expand.grid(BirthWt = c(min(BirthWt), max(BirthWt)),
               Breed = levels(Breed)))

WeanWt.pred <- predict(sheep.lm2, sheep.pred.df)
sheep.pred.df <- cbind(sheep.pred.df, WeanWt.pred)

lines(sheep.pred.df[1:2, -2], col = 1) # Breed 1, DD
lines(sheep.pred.df[3:4, -2], col = 2) # Breed 2, DR
lines(sheep.pred.df[5:6, -2], col = 3) # Breed 3, RD
lines(sheep.pred.df[7:8, -2], col = 4) # Breed 4, RR
abline(h = mean.WW, lty = 2, col = "grey")
abline(v = mean.BW, lty = 2, col = "grey")
```
